# Estimating and Interpreting Effect Sizes {#sec-effect}

## R Setup Used Here

```{r}
#| warning: false
#| message: false

knitr::opts_chunk$set(comment = NA)

library(broom)
library(knitr)
library(rms)
library(tidyverse) 

theme_set(theme_bw())
```

### Data Load

Consider the `smalldat.csv` data available on our site, which is modeled on the public Framingham data set available [from BIOLINCC](https://biolincc.nhlbi.nih.gov/teaching/)[^1]. From the BIOLINCC documentation:

[^1]: The Framingham data from this source are not appropriate for publication or project use because they have been anonymized by permuting the results of individual subjects.

> The Framingham Heart Study is a long term prospective study of the etiology of cardiovascular disease among a population of free living subjects in the community of Framingham, Massachusetts. The Framingham Heart Study was a landmark study in epidemiology in that it was the first prospective study of cardiovascular disease and identified the concept of risk factors and their joint effects.

```{r}
smalldat <- read_csv("data/smalldat.csv", show_col_types = FALSE)
```

## Available Variables

The `smalldat` data contains 150 observations on the following variables[^2]:

[^2]: The *educ* levels are: 1_Low, 2_Middle, 3_High and 4_VHigh, which stands for Very High

| Variable  | Description                                 |
|-----------|---------------------------------------------|
| *subject* | Subject identification code                 |
| *smoker*  | 1 = current smoker, 0 = not current smoker  |
| *totchol* | total cholesterol, in mg/dl                 |
| *age*     | age in years                                |
| *sex*     | subject's sex (M or F)                      |
| *educ*    | subject's educational attainment (4 levels) |

```{r}
ggplot(smalldat, aes(x = totchol)) + 
    geom_histogram(bins = 15, col = "white", fill = "dodgerblue")
```

## Effect Interpretation in A Linear Regression Model

```{r}
m1 <- lm(totchol ~ age + sex + factor(educ),
         data = smalldat)

kable(tidy(m1, conf.int = TRUE), digits = 3)
```

1.  What is the effect of *age* on *totchol* in Model `m1`?

```{r}
temp.a <- tidy(m1, conf.int = TRUE) %>%
    filter(term == "age")

kable(temp.a, digits = 3)
```

The coefficient of the *age* effect on *totchol* is `r (tmp.a <- temp.a %>% select(estimate) %>% round(.,3))`. Suppose we have two subjects, Doris and Emily, who are the same sex and have the same level of education, but Doris is one year older than Emily. Our model predicts that Doris' total cholesterol will be `r tmp.a` mg/dl higher than Emily's.

The 95% confidence interval for this estimated *age* coefficient is (`r (tmp.b <- temp.a %>% select(conf.low) %>% round(.,3))`, `r (tmp.c <- temp.a %>% select(conf.high) %>% round(.,3))`), so holding everything else constant, it seems that older *age* is associated with higher *totchol* in this model.

2.  What is the effect of *sex* on `totchol` in Model `m1`?

```{r}
temp.s <- tidy(m1, conf.int = TRUE) %>%
    filter(term == "sexM")

kable(temp.s, digits = 3)
```

The model is parametrized to incorporate the *sex* information with an indicator (and factor) variable called *sexM* which is interpreted as taking the value 1 when *sex* = M, and 0 otherwise. The coefficient of the *sexM* effect on *totchol* is `r (tmp.s <- temp.s %>% select(estimate) %>% round(.,3))`. Suppose we have two subjects, David and Emily, who are the same age, have the same level of education, but David is male and Emily is female. Our model predicts that David's total cholesterol will be `r tmp.s` mg/dl higher than Emily's.

The 95% confidence interval for this estimated *sexM* coefficient is (`r (tmp.b <- temp.s %>% select(conf.low) %>% round(.,3))`, `r (tmp.c <- temp.s %>% select(conf.high) %>% round(.,3))`), which suggests that the effect of *sex* on *totchol* could be quite small, and that the data are consistent with a wide range of estimates for the *sexM* effect, some of which are negative.

3.  What is the effect of *educ* on *totchol* in Model `m1`?

The *educ* variable splits the subjects into four categories. In this model the "1_Low" category is used as the baseline, and we have estimates for "2_Middle" (as compared to "1_Low"), for "3_High" (as compared to "1_Low") and for "4_VHigh" (as compared to "1_Low".)

```{r}
temp.ed <- tidy(m1, conf.int = TRUE) %>% 
          filter(term %in% c("factor(educ)2_Middle", 
                             "factor(educ)3_High", 
                             "factor(educ)4_VHigh"))

kable(temp.ed, 
      digits = 3)
```

The coefficient of the *educ* effect comparing the "2_Middle" group to the baseline "1_Low" group on *totchol* is `r (tmp.e2 <- temp.ed %>% filter(term == "factor(educ)2_Middle") %>% select(estimate) %>% round(.,3))`.

Note that none of the *educ* levels show especially large differences from the baseline group, and each of their 95% confidence intervals contains zero.

-   Suppose we have two subjects, Lola and Mina, who are the same age, and the same sex, but Lola is in the "1_Low" education group and Mina is in the "2_Middle" education group.
-   Our model predicts that Mina's total cholesterol will be `r tmp.e2` mg/dl higher than Lola's.

The coefficient of the *educ* effect comparing the "3_High" group to the baseline "1_Low" group on *totchol* is `r (tmp.e3 <- temp.ed %>% filter(term == "factor(educ)3_High") %>% select(estimate) %>% round(.,3))`.

-   Suppose we have two subjects, Lola and Heidi, who are the same age, and the same sex, but Lola is in the "1_Low" education group and Heidi is in the "3_High" education group.
-   Our model predicts that Heidi's total cholesterol will be `r abs(tmp.e3)` mg/dl **lower** than Lola's.

Finally, the coefficient of the *educ* effect comparing the "4_VHigh" group to the baseline "1_Low" group on *totchol* is `r (tmp.e4 <- temp.ed %>% filter(term == "factor(educ)4_VHigh") %>% select(estimate) %>% round(.,3))`.

-   Suppose we have two subjects, Lola and Vera, who are the same age, and the same sex, but Lola is in the "1_Low" education group and Vera is in the "4_VHigh" education group.
-   Our model predicts that Vera's total cholesterol will be `r tmp.e4` mg/dl higher than Lola's.

## Making a prediction and building a prediction interval with an `lm` fit

Suppose we want to use `m1` to make a prediction for Lola and Vera, who we'll now assume are each Female and 30 years of age, and we want to accompany this with a 90% prediction interval for each subject. Here's one way to do that.

```{r}
new1 <- tibble(
    name = c("Lola", "Vera"),
    age = c(30, 30),
    sex = c("F", "F"),
    educ = c("1_Low", "4_VHigh")
)

new1

res1 <- predict(m1, newdata = new1, 
        interval = "prediction", level = 0.9)

res1

new1_aug <- bind_cols(new1, fit = res1[,"fit"],
                      pi90.lo = res1[,"lwr"],
                      pi90.hi = res1[,"upr"])

new1_aug
```

## What if we include a Spline or an Interaction?

Suppose we fit a new model to predict *totchol* using a five-knot spline in *age* and the interaction of sex and educational attainment. How does that change our interpretation of the effect sizes?

None of these coefficients show particularly large effects, and zero is contained in each of the 95% confidence intervals provided in the table summarizing model m2.

```{r}
d <- datadist(smalldat); options(datadist = "d")

m2 <- ols(totchol ~ rcs(age, 5) + sex*catg(educ),
         data = smalldat, x = TRUE, y = TRUE)

kable(summary(m2), digits = 2)
```

The `kable` approach I used in these notes hides the adjusted values specified at the bottom of the `summary` table for this `ols` model, but they are `Adjusted to: sex=F educ=1_Low`.

Now, how do we interpret these model `m2` results?

```{r}
plot(summary(m2))
```

-   The **age** interpretation is that if we have two subjects, Al and Bob, who are the same sex and have the same education level, but Al is age `r summary(m2)["age", "Low"]` and Bob is age `r summary(m2)["age", "High"]`, then model `m2` projects that Bob's *totchol* will be `r round(summary(m2)["age", "Effect"],3)` mg/dl higher than will Al's.
-   Because of the interaction between *sex* and *educ* in our model `m2`, we must select an *educ* level in order to cleanly interpret the effect of *sex* on *totchol*. The **sex - M:F** interpretation compares M(ale) to F(emale) *sex* while requiring[^3] that *educ* = `1_Low`. The result is that if we have two subjects, Carl and Diane, who are the same age and *each is in the low education group*, but Carl is Male and Diane is Female, then model `m2` predicts that Carl's *totchol* will be `r round(summary(m2)["sex - M:F", "Effect"],3)` mg/dl higher than will Diane's.
-   Because of the interaction between *sex* and *educ* in our model `m2`, we must select a *sex* in order to cleanly interpret the effect of *educ* on *totchol*. The **educ - 2_Middle:1_Low** term, for instance, compares "2_Middle" education to "1_Low" education while requiring that *sex* is Female[^4]. The result is that if we have two subjects, Lola and Mina, who are the same age and *each is Female*, but Lola is in the "1_Low" education group and Mina is in the "2_Middle" education group, then model `m2` predicts that Mina's *totchol* will be `r round(summary(m2)["educ - 2_Middle:1_Low", "Effect"],3)` mg/dl higher than will Lola's.

[^3]: We know this because of the `Adjusted to: sex = F, educ=1_Low` note at the bottom of the `summary` output for the `ols` model.

[^4]: Again, we know this because of the `Adjusted to: sex = F, educ=1_Low` note.

Here is a nomogram of model `m2`.

```{r}
plot(nomogram(m2))
```

## Making a prediction and building a prediction interval with an `ols` fit

Suppose we want to use `m2` to make a prediction for Lola and Vera, who we'll again assume are each Female and 30 years of age, and we want to accompany this with a 90% prediction interval for each subject. Here's one way to do that.

```{r}
res2_lola <- Predict(m2, 
                     age = 30, sex = "F", educ = "1_Low",
                     conf.int = 0.90,
                     conf.type = "individual")
res2_lola

res2_vera <- Predict(m2, 
                     age = 30, sex = "F", educ = "4_VHigh",
                     conf.int = 0.90,
                     conf.type = "individual")

res2_vera

```


## Effect Estimates in A Logistic Regression fit with `glm`

In a binary **logistic** model, where we predict the log odds of smoking (*smoker* = 1), we will exponentiate so as to interpret the odds ratio estimates associated with each coefficient.

```{r}
m3 <- glm(smoker ~ age + sex + factor(educ), 
         data = smalldat, family = binomial)

kable(tidy(m3, exponentiate = TRUE, conf.int = TRUE), 
      digits = 3)
```

1.  What is the effect of *age* on the odds of being a *smoker* in Model `m3`?

```{r}
temp.3a <- tidy(m3, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "age")

kable(temp.3a, digits = 3)
```

The estimated odds ratio for the *age* effect on *smoker* is `r (tmp.3a <- temp.3a %>% select(estimate) %>% round(.,3))`. Suppose we have two subjects, Doris and Emily, who are the same sex and have the same level of education, but Doris is one year older than Emily. Our model predicts that Doris' odds of smoking will be `r tmp.3a` times as high as Emily's. Another way to write this would be that Doris' odds of smoking are estimated to be `r (100*tmp.3a)`% of Emily's. Yet another way would be to state that Doris' odds of smoking are estimated to be `r (100*(1 - tmp.3a))`% smaller than Emily's odds.

The 95% confidence interval for this estimated odds ratio for the *age* effect on being a *smoker* is (`r (tmp.3b <- temp.3a %>% select(conf.low) %>% round(.,3))`, `r (tmp.3c <- temp.3a %>% select(conf.high) %>% round(.,3))`). This confidence interval for the odds ratio does not include one, and again we see that holding everything else constant, older *age* is associated with detectably lower odds of being a *smoker* in this model.

2.  What is the effect of *sex* on the odds of being a *smoker* in Model `m3`?

```{r}
temp.3s <- tidy(m3, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "sexM")

kable(temp.3s, digits = 3)
```

The model incorporates the *sex* information with an indicator (and factor) variable called *sexM* which is interpreted as taking the value 1 when *sex* = M, and 0 otherwise. The estimated odds ratio describing the *sexM* effect on being a *smoker* is `r (tmp.3s <- temp.3s %>% select(estimate) %>% round(.,3))`. Suppose we have two subjects, David and Emily, who are the same age, have the same level of education, but David is male and Emily is female. Our model predicts that David's odds of being a *smoker* are `r tmp.3s` times the odds that Emily is a *smoker*, or equivalently, that David's odds are `r 100*tmp.3s`% of Emily's odds. Another equivalent statement would be that David's odds are `r (100*tmp.3s - 100)`% larger than Emily's odds.

The 95% confidence interval for the odds ratio estimate of the effect of *sexM* on being a *smoker* is (`r (tmp.b <- temp.3s %>% select(conf.low) %>% round(.,3))`, `r (tmp.c <- temp.3s %>% select(conf.high) %>% round(.,3))`). The effect of *sex* on the odds of being a *smoker* appears modest, and 1 is included in the confidence interval.

3.  What is the effect of *educ* on the odds of being a *smoker* in Model `m3`?

Again, the *educ* variable splits the subjects into four categories. In this model the "1_Low" category is used as the baseline, and we have estimates for "2_Middle" (as compared to "1_Low"), for "3_High" (as compared to "1_Low") and for "4_VHigh" (as compared to "1_Low".)

```{r}
temp.3ed <- tidy(m3, exponentiate = TRUE, conf.int = TRUE) %>% 
          filter(term %in% c("factor(educ)2_Middle", 
                             "factor(educ)3_High", 
                             "factor(educ)4_VHigh"))

kable(temp.3ed, 
      digits = 3)
```

The estimated odds ratio describing the effect of *educ* being "2_Middle" instead of the baseline "1_Low" on the odds of being a *smoker* is `r (tmp.3e2 <- temp.3ed %>% filter(term == "factor(educ)2_Middle") %>% select(estimate) %>% round(.,3))`, for people of the same age and sex.

No *educ* levels show meaningful differences from the baseline group, and their 95% confidence intervals all include 1, although the comparison of 4_VHigh to 1_Low only barely includes 1.

-   Suppose we have two subjects, Lola and Mina, who are the same age, and the same sex, but Lola is in the "1_Low" education group and Mina is in the "2_Middle" education group.
-   Our model predicts that Mina's odds of being a smoker will be `r tmp.3e2` times the odds of Lola being a smoker.

The estimated odds ratio comparing the *educ* = "3_High" group to the baseline *educ* = "1_Low" group on *smoker* is `r (tmp.3e3 <- temp.3ed %>% filter(term == "factor(educ)3_High") %>% select(estimate) %>% round(.,3))`.

-   Suppose we have two subjects, Lola and Heidi, who are the same age, and the same sex, but Lola is in the "1_Low" education group and Heidi is in the "3_High" education group.
-   Our model predicts that Heidi's odds of being a *smoker* will be `r tmp.3e3` times the odds of Lola being a smoker.

Finally, The estimated odds ratio comparing the *educ* = "4_VHigh" group to the baseline *educ* = "1_Low" group on *smoker* is `r (tmp.3e4 <- temp.3ed %>% filter(term == "factor(educ)4_VHigh") %>% select(estimate) %>% round(.,3))`.

-   Suppose we have two subjects, Lola and Vera, who are the same age, and the same sex, but Lola is in the "1_Low" education group and Vera is in the "4_VHigh" education group.
-   Our model predicts that Vera's odds of being a *smoker* will be `r tmp.3e4` times the odds of Lola being a smoker.

## Estimates in The Same Logistic Regression fit with `lrm`

When we fit the same model as `m3` using `lrm`, we get identical results as we get from the `glm` fit for the categorical predictors, but there's a change in how the odds ratio for the quantitative predictor (*age*) is presented.

```{r}
d <- datadist(smalldat); options(datadist = "d")

m3.lrm <- lrm(smoker ~ age + sex + educ, 
         data = smalldat, x = TRUE, y = TRUE)

kable(summary(m3.lrm), digits = 3)
```

Note that the odds ratio effect sizes and confidence intervals are identical to what we saw in the `glm` fit for the *sex* and *educ* variables here, but the *age* result is presented differently.

-   The **age** interpretation is that if we have two subjects, Al and Bob, who are the same sex and have the same education level, but Al is age `r summary(m3.lrm)["age", "Low"]` and Bob is age `r summary(m3.lrm)["age", "High"]`, then model `m3` projects that Bob's odds of being a *smoker* will be `r round(summary(m3.lrm)[2, "Effect"],3)` times Al's odds of being a smoker. Bob's odds are `r round(100 * summary(m3.lrm)[2, "Effect"],1)`% as large as Al's, equivalently.
-   After adjustment for *sex* and *educ*, increasing age appears to be associated with decreasing odds of smoking. Note, too, that the effect of *age* on the odds of being a *smoker* has a confidence interval for the odds ratio entirely below 1.

```{r}
plot(summary(m3.lrm))
```

## Estimates in A New Logistic Regression fit with `lrm`

Now, suppose we fit a new model to predict the log odds of being a *smoker* using a five-knot spline in *age* and the interaction of sex and educational attainment. How does that change our interpretation of the effect sizes?

```{r}
d <- datadist(smalldat); options(datadist = "d")

m4 <- lrm(smoker ~ rcs(age,5) + sex * catg(educ), 
         data = smalldat, x = TRUE, y = TRUE)

kable(summary(m4), digits = 3)
```

Again, the `kable` approach I used in these notes hides the adjusted values specified at the bottom of the `summary` table for this `lrm` model (model `m4`), but they are `Adjusted to: sex=F educ=1_Low`.

Now, how do we interpret these model `m4` results?

```{r}
plot(summary(m4))
```

-   The **age** interpretation is that if we have two subjects, Al and Bob, who are the same sex and have the same education level, but Al is age `r summary(m4)["age", "Low"]` and Bob is age `r summary(m4)["age", "High"]`, then model `m4` projects that Bob's odds of being a smoker will be `r round(summary(m4)[2, "Effect"],3)` times Al's odds of being a smoker. Equivalently, Bob's odds are `r round(100 * summary(m4)[2, "Effect"],1)`% as large as Al's.
-   Because of the interaction between *sex* and *educ* in our model `m4`, we must select an *educ* level in order to cleanly interpret the effect of *sex* on *smoker*. The **sex - M:F** interpretation compares M(ale) to F(emale) *sex* while requiring[^5] that *educ* = `1_Low`. The result is that if we have two subjects, Carl and Diane, who are the same age and *each is in the low education group*, but Carl is Male and Diane is Female, then model `m4` predicts that Carl's odds of being a *smoker* will be `r round(summary(m4)[4, "Effect"],3)` times the odds for Diane.
-   Because of the interaction between *sex* and *educ* in our model `m4`, we must select a *sex* in order to cleanly interpret the effect of *educ* on *totchol*. The **educ - 2_Middle:1_Low** term, for instance, compares "2_Middle" education to "1_Low" education while requiring that *sex* is Female[^6]. The result is that if we have two subjects, Lola and Mina, who are the same age and *each is Female*, but Lola is in the "1_Low" education group and Mina is in the "2_Middle" education group, then model `m4` predicts that Mina's odds of being a *smoker* will be `r round(summary(m4)[6, "Effect"],3)` times Lola's odds, or, equivalently, that Mina will have `r round(100*(summary(m4)[6, "Effect"] - 1),1)`% higher odds than Lola.

[^5]: We know this because of the `Adjusted to: sex = F, educ=1_Low` note at the bottom of the `summary` output.

[^6]: `Adjusted to: sex = F, educ=1_Low` tells us this.

It should be easy to see that one is contained in each of the 95% confidence intervals summarizing model m4 except for the one for the main effect of sex, but we might also consider the impact of the interaction term, as described by the `anova` result for model `m4`.

```{r}
anova(m4)
```

Finally, here is a nomogram of model `m4`.

```{r}
plot(nomogram(m4, fun = plogis))
```
