# R Setup {.unnumbered}

Here, we'll show all of the R packages used in these notes. The list of R Packages we will use in 432 is even more extensive, and is [available on our course website](https://thomaselove.github.io/432-2023/software.html#r-packages-to-install).

```{r}
#| eval: false
#| message: false
#| warning: false

knitr::opts_chunk$set(comment = NA) ## Dr. Love always includes this

library(car)
library(DescTools)
library(Epi)
library(equatiomatic)
library(GGally)
library(ggridges)
library(glue)
library(gt)
library(gtsummary)
library(Hmisc)
library(janitor) 
library(knitr)
library(lmboot)
library(MKinfer)
library(mosaic)
library(naniar)
library(nhanesA)
library(patchwork)
library(rms)
library(rstatix)
library(simputation)
library(vcd)
library(visdat)

library(tidymodels)
library(tidyverse)

```

## General Theme for `ggplot` work {.unnumbered}

Dr. Love prefers `theme_bw()` to the default choice.

```{r}
#| eval: false
theme_set(theme_bw())
```

## Data used in these notes {.unnumbered}

All data sets used in these notes are available on our [432-data website](https://github.com/THOMASELOVE/432-data).

Dr. Love is in the process of moving all of the data loads below to their individual chapters.

```{r}
#| eval: false
#| message: false
#| warning: false

prost <- read_csv("data/prost.csv", show_col_types = FALSE) 

emphysema <- read_csv("data/emphysema.csv", show_col_types = FALSE) 
resect <- read_csv("data/resect.csv", show_col_types = FALSE) 
colscr <- read_csv("data/screening.csv", show_col_types = FALSE) 
colscr2 <- read_csv("data/screening2.csv", show_col_types = FALSE) 
authorship <- read_csv("data/authorship.csv", show_col_types = FALSE) 
hem <- read_csv("data/hem.csv", show_col_types = FALSE) 
leukem <- read_csv("data/leukem.csv", show_col_types = FALSE) 
```
